---
title: "TreeFARMS Technical Details and Implementation"
author: "TreeFARMS Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TreeFARMS Technical Details and Implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Technical Details and Implementation

This vignette provides technical details about the TreeFARMS algorithm, the Rcpp implementation, and the underlying C++ library.

# Algorithm Overview

## TreeFARMS Algorithm

TreeFARMS (Tree-based Fast and Accurate Rule Set Models) is a branch-and-bound algorithm for learning optimal decision trees. The algorithm:

1. **Searches the space of all possible decision trees** using branch-and-bound optimization
2. **Finds trees that minimize a specified loss function** (misclassification or log-loss)
3. **Generates Rashomon sets** of near-optimal trees within a specified bound
4. **Uses advanced pruning techniques** to efficiently explore the search space

## Loss Functions

### Misclassification Loss
The traditional 0-1 loss function:
```
L(y, ŷ) = 1 if y ≠ ŷ, 0 otherwise
```

### Log-Loss (Cross-Entropy)
The probabilistic loss function:
```
L(y, p) = -y*log(p) - (1-y)*log(1-p)
```

## Rashomon Sets

A Rashomon set contains all trees within a specified bound of the optimal solution:
```
R(ε) = {T : L(T) ≤ L(T*) + ε}
```

Where:
- `T*` is the optimal tree
- `ε` is the Rashomon bound
- `L(T)` is the loss of tree T

# Rcpp Implementation

## Architecture

The R package uses direct Rcpp bindings to the C++ implementation:

```
R → Rcpp → C++ GOSDT Library
```

This eliminates the Python dependency and provides:
- **Faster execution** through direct C++ integration
- **Easier distribution** without Python environment setup
- **Better reliability** with fewer dependencies

## Core Rcpp Functions

```{r rcpp-functions}
library(treefarms)

# The main Rcpp functions are:
# - treefarms_configure_cpp(): Configure algorithm parameters
# - treefarms_fit_cpp(): Train model and return JSON results
# - treefarms_fit_and_stats_cpp(): Train model and return statistics
# - treefarms_time_cpp(): Get training time
# - treefarms_iterations_cpp(): Get number of iterations
# - treefarms_size_cpp(): Get model size
# - treefarms_status_cpp(): Get training status

# These functions are called internally by the R wrapper functions
```

## Data Flow

```{r data-flow}
# 1. R data.frame → CSV string
X <- data.frame(x1 = c(0, 1, 0, 1), x2 = c(0, 0, 1, 1))
y <- c(0, 1, 1, 0)

# 2. Combine X and y into CSV format
data_df <- X
data_df$class <- y

# 3. Convert to CSV string
csv_con <- textConnection("csv_data", "w")
write.csv(data_df, csv_con, row.names = FALSE)
close(csv_con)
csv_string <- paste(csv_data, collapse = "\n")

cat("CSV data format:\n")
cat(csv_string)
```

## Configuration

```{r configuration}
# Configuration is passed as JSON
config <- list(
  loss_function = "misclassification",
  regularization = 0.1,
  verbose = FALSE,
  rashomon = TRUE,
  rashomon_bound_multiplier = 0.05
)

config_json <- jsonlite::toJSON(config, auto_unbox = TRUE)
cat("Configuration JSON:\n")
cat(config_json)
```

# C++ Library Details

## GOSDT (Generalized Optimal Sparse Decision Trees)

The underlying C++ library implements the GOSDT algorithm:

- **Branch-and-bound optimization** for finding optimal trees
- **Advanced pruning techniques** for efficient search
- **Parallel processing** using Intel TBB
- **High-precision arithmetic** using GMP for numerical stability

## Dependencies

### Intel TBB (Threading Building Blocks)
- Provides parallel processing capabilities
- Used for multi-threaded tree search
- Required for optimal performance

### GMP (GNU Multiple Precision)
- Provides high-precision arithmetic
- Ensures numerical stability in optimization
- Required for accurate loss calculations

## Build System

The package uses standard Rcpp build system:

```makefile
# src/Makevars
PKG_CPPFLAGS = -I../inst/include -I.
PKG_LIBS = -ltbb -ltbbmalloc -lgmp -lpthread
CXX_STD = CXX11
```

# Performance Characteristics

## Time Complexity

The algorithm has exponential worst-case complexity, but practical performance is much better due to:

1. **Advanced pruning techniques** that eliminate large portions of the search space
2. **Parallel processing** that distributes work across multiple cores
3. **Efficient data structures** for fast tree evaluation

## Memory Usage

Memory usage scales with:
- **Dataset size**: O(n) for n observations
- **Number of features**: O(2^p) in worst case for p features
- **Rashomon set size**: O(k) for k trees in the set

## Scalability

```{r scalability-test}
library(treefarms)

# Test scalability with different dataset sizes
dataset_sizes <- c(100, 500, 1000, 2000)
results <- data.frame(
  n_obs = dataset_sizes,
  time_seconds = numeric(length(dataset_sizes)),
  memory_mb = numeric(length(dataset_sizes)),
  n_trees = numeric(length(dataset_sizes))
)

for (i in seq_along(dataset_sizes)) {
  n <- dataset_sizes[i]
  
  # Create dataset
  X <- data.frame(
    feature_1 = sample(0:1, n, replace = TRUE),
    feature_2 = sample(0:1, n, replace = TRUE),
    feature_3 = sample(0:1, n, replace = TRUE)
  )
  y <- sample(0:1, n, replace = TRUE)
  
  # Measure performance
  start_time <- Sys.time()
  mem_before <- pryr::mem_used()
  
  model <- treefarms(X, y, 
                    loss_function = "misclassification", 
                    regularization = 0.01,
                    verbose = FALSE)
  
  end_time <- Sys.time()
  mem_after <- pryr::mem_used()
  
  results$time_seconds[i] <- as.numeric(end_time - start_time, units = "secs")
  results$memory_mb[i] <- (mem_after - mem_before) / 1024^2
  results$n_trees[i] <- model$n_trees
}

print(results)
```

# Error Handling and Debugging

## Common Error Conditions

### Compilation Errors
```{r compilation-errors}
# Missing system dependencies
# Error: fatal error: 'tbb/tbb.h' file not found
# Solution: Install Intel TBB

# Error: fatal error: 'gmp.h' file not found  
# Solution: Install GMP library

# Error: symbol not found: '_scalable_free'
# Solution: Link against libtbbmalloc
```

### Runtime Errors
```{r runtime-errors}
# Invalid input data
X_invalid <- data.frame(x1 = c(0, 1, 2))  # Non-binary values
y_invalid <- c(0, 1, 2)  # Non-binary values

# These will cause errors:
# treefarms(X_invalid, y_invalid)  # Error: Feature must contain only binary values
# treefarms(X, y_invalid)  # Error: y must contain only binary values
```

## Debugging Tips

### Verbose Output
```{r verbose-debugging}
# Use verbose = TRUE for debugging
model <- treefarms(X, y, 
                  loss_function = "misclassification", 
                  regularization = 0.01,
                  verbose = TRUE)
```

### Check Model Status
```{r model-status}
# Check training status
cat("Training time:", model$training_time, "seconds\n")
cat("Training iterations:", model$training_iterations, "\n")
cat("Model size:", model$model$size, "\n")
cat("Training status:", model$model$status, "\n")
```

# Advanced Configuration

## Custom Parameters

```{r custom-parameters}
# Advanced configuration options
config_advanced <- list(
  loss_function = "log_loss",
  regularization = 0.01,
  verbose = TRUE,
  rashomon = TRUE,
  rashomon_bound_multiplier = 0.05,
  target_trees = 3,
  max_trees = 10,
  # Additional C++ parameters can be added here
  time_limit = 300,  # 5 minutes
  memory_limit = 1000  # 1GB
)

# Note: Some parameters may not be exposed in the R interface
# Check the C++ documentation for full parameter list
```

## Memory Management

```{r memory-management}
# Monitor memory usage
library(pryr)

# Check memory before training
mem_before <- mem_used()
cat("Memory before:", round(mem_before / 1024^2, 2), "MB\n")

# Train model
model <- treefarms(X, y, 
                  loss_function = "misclassification", 
                  regularization = 0.01,
                  verbose = FALSE)

# Check memory after training
mem_after <- mem_used()
mem_increase <- mem_after - mem_before
cat("Memory increase:", round(mem_increase / 1024^2, 2), "MB\n")

# Clean up if needed
rm(model)
gc()
```

# Integration with Other R Packages

## Parallel Processing

```{r parallel-processing}
# Use parallel processing for multiple models
library(parallel)

# Create cluster
cl <- makeCluster(4)

# Train multiple models in parallel
regularization_values <- c(0.001, 0.01, 0.1, 0.5)
models <- parLapply(cl, regularization_values, function(reg) {
  treefarms(X, y, 
            loss_function = "misclassification", 
            regularization = reg,
            verbose = FALSE)
})

# Stop cluster
stopCluster(cl)

# Compare results
results <- data.frame(
  regularization = regularization_values,
  n_trees = sapply(models, function(m) m$n_trees),
  accuracy = sapply(models, function(m) m$accuracy)
)

print(results)
```

## Integration with caret

```{r caret-integration}
# Example of integrating with caret (if available)
# library(caret)

# Define custom model
# treefarms_model <- list(
#   label = "TreeFARMS",
#   library = "treefarms",
#   type = "Classification",
#   parameters = data.frame(
#     parameter = c("regularization", "loss_function"),
#     class = c("numeric", "character"),
#     label = c("Regularization", "Loss Function")
#   ),
#   grid = function(x, y, len = NULL) {
#     expand.grid(
#       regularization = c(0.001, 0.01, 0.1, 0.5),
#       loss_function = c("misclassification", "log_loss")
#     )
#   },
#   fit = function(x, y, wts, param, lev, last, classProbs, ...) {
#     treefarms(x, y, 
#               regularization = param$regularization,
#               loss_function = param$loss_function,
#               ...)
#   },
#   predict = function(modelFit, newdata, submodels = NULL) {
#     predict(modelFit, newdata, type = "class")
#   },
#   prob = function(modelFit, newdata, submodels = NULL) {
#     predict(modelFit, newdata, type = "prob")
#   }
# )
```

# Conclusion

The TreeFARMS R package provides a robust, high-performance implementation of the TreeFARMS algorithm through direct Rcpp bindings to the C++ library. Key technical advantages include:

1. **Direct C++ integration** for optimal performance
2. **Advanced branch-and-bound optimization** for finding optimal trees
3. **Parallel processing** using Intel TBB
4. **High-precision arithmetic** using GMP
5. **Comprehensive error handling** and debugging support
6. **Standard R package structure** for easy integration

The implementation balances theoretical rigor with practical performance, making it suitable for both research and production applications.


