---
title: "TreeFARMS Advanced Usage and Performance"
author: "TreeFARMS Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TreeFARMS Advanced Usage and Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Advanced TreeFARMS Usage

This vignette covers advanced usage patterns, performance optimization, and real-world applications of the TreeFARMS R package.

# Performance Optimization

## Large Dataset Handling

```{r large-dataset}
library(treefarms)

# Create larger dataset for performance testing
set.seed(42)
n <- 1000
X_large <- data.frame(
  feature_1 = sample(0:1, n, replace = TRUE),
  feature_2 = sample(0:1, n, replace = TRUE),
  feature_3 = sample(0:1, n, replace = TRUE),
  feature_4 = sample(0:1, n, replace = TRUE),
  feature_5 = sample(0:1, n, replace = TRUE)
)

# Create target with complex pattern
y_large <- as.numeric((X_large$feature_1 == 1 & X_large$feature_2 == 1) | 
                      (X_large$feature_3 == 1 & X_large$feature_4 == 1))

# Time the training
start_time <- Sys.time()
model_large <- treefarms(X_large, y_large, 
                        loss_function = "misclassification", 
                        regularization = 0.01,
                        verbose = FALSE)
end_time <- Sys.time()

cat("Training time:", round(as.numeric(end_time - start_time, units = "secs"), 3), "seconds\n")
cat("Number of trees found:", model_large$n_trees, "\n")
cat("Training accuracy:", round(model_large$accuracy, 3), "\n")
```

## Memory Usage Optimization

```{r memory-optimization}
# Monitor memory usage
library(pryr)

# Check memory before training
mem_before <- mem_used()

# Train model
model_mem <- treefarms(X_large, y_large, 
                      loss_function = "misclassification", 
                      regularization = 0.01,
                      verbose = FALSE)

# Check memory after training
mem_after <- mem_used()
mem_increase <- mem_after - mem_before

cat("Memory increase:", round(mem_increase / 1024^2, 2), "MB\n")
cat("Model size:", round(object_size(model_mem) / 1024^2, 2), "MB\n")
```

# Advanced Cross-Fitted Rashomon Analysis

## Multi-Fold Analysis

```{r multi-fold-analysis}
# Analyze different K values
K_values <- c(2, 3, 5, 10)
results <- list()

for (K in K_values) {
  cat("Analyzing K =", K, "...\n")
  result <- cross_fitted_rashomon(X_large, y_large, 
                                 K = K,
                                 loss_function = "misclassification", 
                                 regularization = 0.01,
                                 verbose = FALSE)
  results[[as.character(K)]] <- result
}

# Compare results
comparison_df <- data.frame(
  K = K_values,
  n_intersecting = sapply(results, function(r) r$n_intersecting),
  avg_rashomon_size = sapply(results, function(r) mean(r$rashomon_sizes)),
  min_rashomon_size = sapply(results, function(r) min(r$rashomon_sizes)),
  max_rashomon_size = sapply(results, function(r) max(r$rashomon_sizes))
)

print(comparison_df)
```

## Stability Analysis

```{r stability-analysis}
# Analyze stability across different random seeds
seeds <- c(42, 123, 456, 789, 999)
stability_results <- list()

for (seed in seeds) {
  set.seed(seed)
  result <- cross_fitted_rashomon(X_large, y_large, 
                                 K = 5,
                                 loss_function = "misclassification", 
                                 regularization = 0.01,
                                 seed = seed,
                                 verbose = FALSE)
  stability_results[[as.character(seed)]] <- result
}

# Analyze stability
stability_df <- data.frame(
  seed = seeds,
  n_intersecting = sapply(stability_results, function(r) r$n_intersecting),
  avg_rashomon_size = sapply(stability_results, function(r) mean(r$rashomon_sizes))
)

print(stability_df)

# Calculate stability metrics
cat("Mean intersecting trees:", round(mean(stability_df$n_intersecting), 2), "\n")
cat("SD of intersecting trees:", round(sd(stability_df$n_intersecting), 2), "\n")
cat("Mean Rashomon size:", round(mean(stability_df$avg_rashomon_size), 2), "\n")
```

# Real-World Applications

## Medical Diagnosis Example

```{r medical-example}
# Simulate medical diagnosis data
set.seed(42)
n_patients <- 500

# Patient features (binary symptoms)
patients <- data.frame(
  fever = sample(0:1, n_patients, replace = TRUE, prob = c(0.7, 0.3)),
  cough = sample(0:1, n_patients, replace = TRUE, prob = c(0.6, 0.4)),
  headache = sample(0:1, n_patients, replace = TRUE, prob = c(0.8, 0.2)),
  fatigue = sample(0:1, n_patients, replace = TRUE, prob = c(0.7, 0.3)),
  nausea = sample(0:1, n_patients, replace = TRUE, prob = c(0.9, 0.1))
)

# Disease diagnosis (1 = disease, 0 = healthy)
# Complex rule: (fever AND cough) OR (headache AND fatigue AND nausea)
disease <- as.numeric((patients$fever == 1 & patients$cough == 1) | 
                      (patients$headache == 1 & patients$fatigue == 1 & patients$nausea == 1))

# Train model
medical_model <- treefarms(patients, disease, 
                          loss_function = "log_loss", 
                          regularization = 0.01,
                          verbose = FALSE)

cat("Medical diagnosis model:\n")
cat("Number of trees:", medical_model$n_trees, "\n")
cat("Training accuracy:", round(medical_model$accuracy, 3), "\n")

# Get probability predictions
probabilities <- medical_model$probabilities
disease_prob <- probabilities[, 2]

# Analyze probability distribution
cat("Disease probability range:", round(range(disease_prob), 3), "\n")
cat("Mean disease probability:", round(mean(disease_prob), 3), "\n")
```

## Financial Risk Assessment

```{r financial-example}
# Simulate financial risk data
set.seed(42)
n_loans <- 800

# Loan applicant features
applicants <- data.frame(
  high_income = sample(0:1, n_loans, replace = TRUE, prob = c(0.6, 0.4)),
  good_credit = sample(0:1, n_loans, replace = TRUE, prob = c(0.7, 0.3)),
  stable_employment = sample(0:1, n_loans, replace = TRUE, prob = c(0.8, 0.2)),
  low_debt_ratio = sample(0:1, n_loans, replace = TRUE, prob = c(0.6, 0.4)),
  collateral = sample(0:1, n_loans, replace = TRUE, prob = c(0.7, 0.3))
)

# Default risk (1 = default, 0 = no default)
# Rule: (low_income AND high_debt) OR (bad_credit AND no_collateral)
default <- as.numeric((applicants$high_income == 0 & applicants$low_debt_ratio == 0) | 
                      (applicants$good_credit == 0 & applicants$collateral == 0))

# Train model
risk_model <- treefarms(applicants, default, 
                       loss_function = "log_loss", 
                       regularization = 0.01,
                       verbose = FALSE)

cat("Financial risk model:\n")
cat("Number of trees:", risk_model$n_trees, "\n")
cat("Training accuracy:", round(risk_model$accuracy, 3), "\n")

# Cross-fitted Rashomon for stability
risk_cf <- cross_fitted_rashomon(applicants, default, 
                                K = 5,
                                loss_function = "log_loss", 
                                regularization = 0.01,
                                verbose = FALSE)

cat("Cross-fitted Rashomon results:\n")
cat("Intersecting trees:", risk_cf$n_intersecting, "\n")
cat("Rashomon sizes:", paste(risk_cf$rashomon_sizes, collapse = ", "), "\n")
```

# Model Interpretation and Visualization

## Tree Rule Extraction

```{r tree-rules}
# Extract and analyze tree rules
trees <- get_rashomon_trees(medical_model)

if (length(trees) > 0) {
  cat("Extracting rules from medical diagnosis model:\n")
  
  for (i in seq_along(trees)) {
    cat("\n--- Tree", i, "---\n")
    rules <- get_tree_rules(trees[[i]], 
                           feature_names = c("fever", "cough", "headache", "fatigue", "nausea"))
    cat(paste(rules, collapse = "\n"), "\n")
  }
}
```

## Probability Calibration Analysis

```{r probability-calibration}
# Analyze probability calibration
probabilities <- medical_model$probabilities
disease_prob <- probabilities[, 2]

# Create probability bins
bins <- cut(disease_prob, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)
bin_centers <- seq(0.05, 0.95, by = 0.1)

# Calculate observed vs predicted probabilities
calibration_data <- data.frame(
  bin = bin_centers,
  predicted = bin_centers,
  observed = tapply(disease, bins, mean),
  count = tapply(disease, bins, length)
)

# Remove bins with no data
calibration_data <- calibration_data[!is.na(calibration_data$observed), ]

print(calibration_data)

# Calculate calibration error
calibration_error <- mean(abs(calibration_data$predicted - calibration_data$observed))
cat("Mean calibration error:", round(calibration_error, 3), "\n")
```

# Performance Benchmarking

## Speed Comparison

```{r speed-comparison}
# Compare different regularization values
reg_values <- c(0.001, 0.01, 0.1, 0.5)
timing_results <- data.frame(
  regularization = reg_values,
  time_seconds = numeric(length(reg_values)),
  n_trees = numeric(length(reg_values)),
  accuracy = numeric(length(reg_values))
)

for (i in seq_along(reg_values)) {
  reg <- reg_values[i]
  
  start_time <- Sys.time()
  model <- treefarms(X_large, y_large, 
                    loss_function = "misclassification", 
                    regularization = reg,
                    verbose = FALSE)
  end_time <- Sys.time()
  
  timing_results$time_seconds[i] <- as.numeric(end_time - start_time, units = "secs")
  timing_results$n_trees[i] <- model$n_trees
  timing_results$accuracy[i] <- model$accuracy
}

print(timing_results)
```

## Memory Usage Analysis

```{r memory-analysis}
# Analyze memory usage across different dataset sizes
dataset_sizes <- c(100, 500, 1000, 2000)
memory_results <- data.frame(
  n_obs = dataset_sizes,
  memory_mb = numeric(length(dataset_sizes)),
  n_trees = numeric(length(dataset_sizes))
)

for (i in seq_along(dataset_sizes)) {
  n <- dataset_sizes[i]
  
  # Create dataset
  X_test <- data.frame(
    feature_1 = sample(0:1, n, replace = TRUE),
    feature_2 = sample(0:1, n, replace = TRUE),
    feature_3 = sample(0:1, n, replace = TRUE)
  )
  y_test <- sample(0:1, n, replace = TRUE)
  
  # Measure memory
  mem_before <- mem_used()
  model <- treefarms(X_test, y_test, 
                    loss_function = "misclassification", 
                    regularization = 0.01,
                    verbose = FALSE)
  mem_after <- mem_used()
  
  memory_results$memory_mb[i] <- (mem_after - mem_before) / 1024^2
  memory_results$n_trees[i] <- model$n_trees
}

print(memory_results)
```

# Best Practices for Production

## Model Validation

```{r model-validation}
# Create train/test split
set.seed(42)
train_idx <- sample(1:nrow(X_large), 0.7 * nrow(X_large))
X_train <- X_large[train_idx, ]
y_train <- y_large[train_idx]
X_test <- X_large[-train_idx, ]
y_test <- y_large[-train_idx]

# Train model
model_val <- treefarms(X_train, y_train, 
                      loss_function = "misclassification", 
                      regularization = 0.01,
                      verbose = FALSE)

# Test predictions
test_pred <- predict(model_val, X_test, type = "class")
test_prob <- predict(model_val, X_test, type = "prob")

# Calculate test accuracy
test_accuracy <- mean(test_pred == y_test)
cat("Training accuracy:", round(model_val$accuracy, 3), "\n")
cat("Test accuracy:", round(test_accuracy, 3), "\n")

# Calculate test log-loss
test_logloss <- -mean(y_test * log(test_prob[, 2] + 1e-15) + 
                      (1 - y_test) * log(test_prob[, 1] + 1e-15))
cat("Test log-loss:", round(test_logloss, 3), "\n")
```

## Cross-Validation for Parameter Selection

```{r cv-parameter-selection}
# Cross-validation for regularization selection
reg_values <- c(0.001, 0.01, 0.1, 0.5)
cv_results <- data.frame(
  regularization = reg_values,
  cv_accuracy = numeric(length(reg_values)),
  cv_logloss = numeric(length(reg_values))
)

K_cv <- 5
fold_size <- nrow(X_train) %/% K_cv

for (i in seq_along(reg_values)) {
  reg <- reg_values[i]
  cv_accuracies <- numeric(K_cv)
  cv_loglosses <- numeric(K_cv)
  
  for (k in 1:K_cv) {
    # Create fold
    start_idx <- (k - 1) * fold_size + 1
    end_idx <- min(k * fold_size, nrow(X_train))
    val_idx <- start_idx:end_idx
    
    X_cv_train <- X_train[-val_idx, ]
    y_cv_train <- y_train[-val_idx]
    X_cv_val <- X_train[val_idx, ]
    y_cv_val <- y_train[val_idx]
    
    # Train model
    model_cv <- treefarms(X_cv_train, y_cv_train, 
                         loss_function = "log_loss", 
                         regularization = reg,
                         verbose = FALSE)
    
    # Validate
    val_pred <- predict(model_cv, X_cv_val, type = "class")
    val_prob <- predict(model_cv, X_cv_val, type = "prob")
    
    cv_accuracies[k] <- mean(val_pred == y_cv_val)
    cv_loglosses[k] <- -mean(y_cv_val * log(val_prob[, 2] + 1e-15) + 
                             (1 - y_cv_val) * log(val_prob[, 1] + 1e-15))
  }
  
  cv_results$cv_accuracy[i] <- mean(cv_accuracies)
  cv_results$cv_logloss[i] <- mean(cv_loglosses)
}

print(cv_results)

# Select best regularization
best_reg_idx <- which.min(cv_results$cv_logloss)
best_reg <- cv_results$regularization[best_reg_idx]
cat("Best regularization:", best_reg, "\n")
cat("Best CV log-loss:", round(cv_results$cv_logloss[best_reg_idx], 3), "\n")
```

# Conclusion

This vignette demonstrated advanced usage patterns for the TreeFARMS R package, including:

1. **Performance optimization** for large datasets
2. **Advanced cross-fitted Rashomon analysis** with stability assessment
3. **Real-world applications** in medical diagnosis and financial risk
4. **Model interpretation** through rule extraction and probability calibration
5. **Performance benchmarking** and memory usage analysis
6. **Production best practices** including validation and parameter selection

The TreeFARMS R package provides a robust foundation for building interpretable decision tree models with strong theoretical guarantees and practical performance characteristics.


