---
title: "TreeFARMS R Package: Introduction and Examples"
author: "TreeFARMS Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TreeFARMS R Package: Introduction and Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The TreeFARMS R package provides a comprehensive interface to TreeFARMS (Tree-based Fast and Accurate Rule Set Models), a powerful algorithm for learning optimal decision trees. This package uses direct Rcpp bindings to the C++ implementation, eliminating Python dependencies and providing:

- **Fast execution** with direct C++ integration
- **Log-loss optimization** for probabilistic modeling
- **Cross-fitted Rashomon sets** for finding stable decision trees
- **Well-calibrated probability predictions** bounded away from 0 and 1
- **Automatic parameter tuning** for optimal model selection

## Key Features

- ✅ **Direct Rcpp bindings**: No Python dependencies, faster execution
- ✅ **Log-loss support**: Cross-entropy loss for probabilistic modeling
- ✅ **Cross-fitted Rashomon sets**: Find trees stable across multiple folds
- ✅ **Auto-tuning**: Automatic parameter optimization
- ✅ **Comprehensive S3 methods**: Standard R interface
- ✅ **Binary feature support**: Optimized for binary decision trees

# Installation

## Prerequisites

1. **System Dependencies**: Install required system libraries

**macOS:**
```bash
brew install tbb gmp
```

**Ubuntu/Debian:**
```bash
sudo apt-get install libtbb-dev libgmp-dev
```

**Windows:**
- Install Rtools (includes GMP)
- Install Intel TBB from https://github.com/oneapi-src/oneTBB/releases

2. **R Package Dependencies**: Install required R packages

```{r install-deps, eval = FALSE}
install.packages(c("Rcpp", "jsonlite", "devtools"))
```

## Install TreeFARMS R Package

```{r install-package, eval = FALSE}
# Install from source
devtools::install("path/to/treefarms")

# Or if using roxygen2 for documentation
devtools::document()
devtools::install()
```

# Quick Start

## Basic Usage

```{r basic-usage}
library(treefarms)

# Create sample binary data with clear pattern
set.seed(42)
n <- 100
X <- data.frame(
  x1 = c(rep(0, 25), rep(1, 25), rep(0, 25), rep(1, 25)),
  x2 = c(rep(0, 25), rep(0, 25), rep(1, 25), rep(1, 25))
)

# Create target with XOR pattern
y <- c(rep(0, 50), rep(1, 50))  # x1 XOR x2

# Train with misclassification loss
model <- treefarms(X, y, loss_function = "misclassification", regularization = 0.01)

# View model summary
print(model)
```

## Loss Function Options

### Misclassification Loss

```{r misclassification-example}
# Traditional misclassification loss
model_misclass <- treefarms(X, y, 
                           loss_function = "misclassification",
                           regularization = 0.01)

# Get predictions
predictions <- model_misclass$predictions
probabilities <- model_misclass$probabilities

cat("Number of trees found:", model_misclass$n_trees, "\n")
cat("Training accuracy:", round(model_misclass$accuracy, 3), "\n")
head(probabilities)
```

### Log-Loss (Cross-Entropy)

```{r log-loss-example}
# Log-loss for probabilistic modeling
model_logloss <- treefarms(X, y, 
                          loss_function = "log_loss", 
                          regularization = 0.01)

# Get probability predictions
probabilities <- model_logloss$probabilities
head(probabilities)

# Compare probability ranges
cat("Log-loss probability range:", 
    paste(round(range(model_logloss$probabilities), 3), collapse = " - "), "\n")
cat("Misclassification probability range:", 
    paste(round(range(model_misclass$probabilities), 3), collapse = " - "), "\n")
```

# Cross-Fitted Rashomon Sets (Primary Feature)

The cross-fitted Rashomon functionality is the primary feature of this package, designed to find stable decision trees that appear across multiple cross-fitting folds.

## Basic Cross-Fitted Rashomon

```{r cross-fitted-basic}
# Find stable trees across cross-fitting folds
result <- cross_fitted_rashomon(X, y, 
                               K = 3, 
                               loss_function = "misclassification", 
                               regularization = 0.01)

# Check results
print(result)
```

## Understanding the Results

```{r cross-fitted-details}
# Examine the results in detail
cat("Number of folds:", result$K, "\n")
cat("Loss function:", result$loss_function, "\n")
cat("Regularization:", result$regularization, "\n")
cat("Trees per fold:", paste(result$rashomon_sizes, collapse = ", "), "\n")
cat("Intersecting trees:", result$n_intersecting, "\n")

if (result$n_intersecting > 0) {
  cat("✓ Found stable tree(s) appearing in all folds!\n")
} else {
  cat("✗ No trees found in all folds.\n")
}
```

## Making Predictions with Stable Trees

```{r cross-fitted-predictions}
# Create new data for prediction
X_new <- data.frame(
  x1 = c(0, 1, 0, 1),
  x2 = c(0, 0, 1, 1)
)

if (result$n_intersecting > 0) {
  # Get class predictions
  pred_class <- predict(result, X_new, type = "class")
  cat("Class predictions:", pred_class, "\n")
  
  # Get probability predictions
  pred_prob <- predict(result, X_new, type = "prob")
  print("Probability predictions:")
  print(pred_prob)
}
```

# Auto-Tuning Parameters

The package includes automatic parameter tuning functionality, allowing users to specify only one parameter and automatically find the optimal value for the other.

## Auto-Tune Regularization

```{r auto-tune-regularization}
# Auto-tune regularization (fix rashomon_bound_multiplier)
model_auto <- treefarms(X, y, 
                       loss_function = "misclassification",
                       regularization = NULL,  # Auto-tune this
                       rashomon_bound_multiplier = 0.05,  # Fix this
                       target_trees = 1,
                       verbose = TRUE)

cat("Auto-tuned regularization:", model_auto$regularization, "\n")
cat("Trees found:", model_auto$n_trees, "\n")
```

## Auto-Tune Rashomon Bound Multiplier

```{r auto-tune-rashomon}
# Auto-tune rashomon_bound_multiplier (fix regularization)
model_auto2 <- treefarms(X, y, 
                        loss_function = "misclassification",
                        regularization = 0.01,  # Fix this
                        rashomon_bound_multiplier = NULL,  # Auto-tune this
                        target_trees = 2,
                        verbose = TRUE)

cat("Auto-tuned rashomon_bound_multiplier:", model_auto2$rashomon_bound_multiplier, "\n")
cat("Trees found:", model_auto2$n_trees, "\n")
```

# Advanced Usage

## Working with Rashomon Sets

```{r rashomon-utils}
# Get individual trees from Rashomon set
trees <- get_rashomon_trees(model_logloss)
cat("Number of trees in Rashomon set:", length(trees), "\n")

if (length(trees) > 0) {
  # Convert tree to JSON
  tree_json <- tree_to_json(trees[[1]])
  cat("First tree JSON length:", nchar(tree_json), "\n")
  
  # Get human-readable rules
  rules <- get_tree_rules(trees[[1]], feature_names = c("x1", "x2"))
  cat("Tree rules:\n")
  cat(paste(rules, collapse = "\n"), "\n")
}
```

## Model Comparison

```{r model-comparison}
# Compare different models
models <- list(
  "Misclassification (reg=0.001)" = treefarms(X, y, loss_function = "misclassification", regularization = 0.001),
  "Misclassification (reg=0.01)" = treefarms(X, y, loss_function = "misclassification", regularization = 0.01),
  "Misclassification (reg=0.1)" = treefarms(X, y, loss_function = "misclassification", regularization = 0.1)
)

# Compare results
comparison <- data.frame(
  Model = names(models),
  Trees = sapply(models, function(m) m$n_trees),
  Accuracy = sapply(models, function(m) round(m$accuracy, 3)),
  Regularization = sapply(models, function(m) m$regularization)
)

print(comparison)
```

## Prediction with New Data

```{r prediction-example}
# Create new data for prediction
X_new <- data.frame(
  x1 = c(0, 1, 0, 1, 0, 1),
  x2 = c(0, 0, 1, 1, 0, 1)
)

# Get predictions from trained model
pred_class <- predict(model_misclass, X_new, type = "class")
pred_prob <- predict(model_misclass, X_new, type = "prob")

# Display results
results_df <- data.frame(
  x1 = X_new$x1,
  x2 = X_new$x2,
  predicted_class = pred_class,
  prob_class_0 = round(pred_prob[, 1], 3),
  prob_class_1 = round(pred_prob[, 2], 3)
)

print(results_df)
```

# Best Practices

## Parameter Selection

1. **Misclassification**: Start with regularization = 0.01, adjust based on tree count
2. **Log-loss**: Start with regularization = 0.01, adjust based on performance
3. **Cross-fitted Rashomon**: Use K = 3-5 for good balance of stability and computational cost

## Data Requirements

- **Binary features only**: All features must be 0 or 1
- **Binary targets**: Target variable must be 0 or 1
- **Sufficient data**: At least 50-100 observations recommended
- **Clear patterns**: Algorithm works best with interpretable decision boundaries

## Performance Tips

- Use `verbose = TRUE` for debugging
- Start with lower regularization for more trees
- Use auto-tuning for parameter optimization
- Check cross-fitted Rashomon results for model stability
- Use clear patterns in data for better tree generation

# Troubleshooting

## Common Issues

### No Trees Generated
```{r troubleshooting-no-trees}
# If no trees are generated, try lower regularization
model_no_trees <- treefarms(X, y, loss_function = "misclassification", regularization = 0.1)
cat("Trees with reg=0.1:", model_no_trees$n_trees, "\n")

model_with_trees <- treefarms(X, y, loss_function = "misclassification", regularization = 0.01)
cat("Trees with reg=0.01:", model_with_trees$n_trees, "\n")
```

### No Intersecting Trees in Cross-Fitted Rashomon
```{r troubleshooting-no-intersection}
# If no intersecting trees, try different parameters
result_no_intersection <- cross_fitted_rashomon(X, y, K = 3, 
                                               loss_function = "misclassification", 
                                               regularization = 0.1)
cat("Intersecting trees with reg=0.1:", result_no_intersection$n_intersecting, "\n")

result_with_intersection <- cross_fitted_rashomon(X, y, K = 3, 
                                                 loss_function = "misclassification", 
                                                 regularization = 0.01)
cat("Intersecting trees with reg=0.01:", result_with_intersection$n_intersecting, "\n")
```

### Auto-tuning Not Finding Target Trees
```{r troubleshooting-auto-tuning}
# If auto-tuning doesn't find target trees, adjust target
model_auto_low_target <- treefarms(X, y, 
                                  loss_function = "misclassification",
                                  regularization = NULL,
                                  target_trees = 1,  # Lower target
                                  verbose = TRUE)

cat("Auto-tuned with target=1, trees found:", model_auto_low_target$n_trees, "\n")
```

## Getting Help

- Check the package documentation: `help(package = "treefarms")`
- View function help: `?treefarms`, `?cross_fitted_rashomon`
- Check the GitHub repository for issues and examples

# Conclusion

The TreeFARMS R package provides a powerful and user-friendly interface to TreeFARMS, with particular strength in:

1. **Direct Rcpp integration** for fast execution without Python dependencies
2. **Log-loss optimization** for probabilistic modeling
3. **Cross-fitted Rashomon sets** for finding stable decision trees
4. **Automatic parameter tuning** for optimal model selection
5. **Comprehensive S3 methods** for standard R integration

The package is designed to be both powerful for advanced users and accessible for beginners, with automatic parameter tuning and comprehensive error handling making it easy to get started while providing the flexibility needed for complex modeling tasks.

## Key Advantages of Rcpp Implementation

- **Faster execution**: Direct C++ integration eliminates Python overhead
- **Easier distribution**: No Python environment setup required
- **Better reliability**: Fewer dependencies and potential failure points
- **Standard R package**: Follows typical Rcpp package patterns
- **Easier maintenance**: One less layer to debug and maintain